A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2024-03-25 docs/manuscript.Rmd
add marginaleffects code on interval estimation
initial work on estimand intro section

- Lakens ebook
- R for data science
- Bruine article, faux documentation
- draschkow power tutorial
- what is your estimand? paper

@@

@@

The first step is to write code that simulates the (assumed) data generating process of the planned study.
This requires specifying a generative process for all predictor variables used in the final data analysis.
While such assumptions can be quite challenging for observational studies and/or continuous predictor variables, this is less of a problem for experimental studies with categorical predictor variables.
When all predictor variables have been simulated, one can use the structure of a suitable GLMM to simulate the dependent variable.
To simulate the GLMM, one requires plausible values for all model parameters.
We will discuss strategies for how these values can be obtained later.
Because one has full control over the data generating process in a tailored simulation, it is simple to also model highly specific aspects of the planned study, like data missing by design or assuming that subjects dropout (i.e. deleting their data from the simulated dataset) based on other variables in the simulated dataset.
The quality of the results of the sample size planning crucially depends on the plausibility of the simulated data generating process.
For this reasons, deciding on a specific set of assumptions can easily feel overwelming.
However, we would argue that even a strongly simplified data generating process (e.g. including only a small number of interaction effect; including only random intercepts but no random slopes; making rather implausible assumptions for missing data) can yield informative result and should be strongly preferred to the current state of psychological research where most studies do not justify their sample size or use highly questionable heuristics that are independent of their specific research question.

@@

### Specify the estimand

This step requires a clear definition of which theoretical quantity is necessary to answer the specific research question of the study under planning.
In an ideal world, this should be the first step in every research process.
However in our experience, many psychological researchers are still not well trained in this task.
We think that simulating data from a generic GLMM that is suited to describe (at least a rough approximation of) the assumed data generating process, can help to better reason about what information is actually needed to answer the specific research question, and how this information is actually represented in the context of the statistical model representing the data generating process of the planned study.

@@

## Strategies on specifying the population parameters

### Insightful descriptive statistics

### Insightful model based quantities

### Iterative process with domain experts

@@

>>>>

- Lakens ebook evtl. Andy Field für die Basics?
- R for data science, Bruine article, faux documentation für die R skills
- kumle power tutorial und irgendwelche paper (am besten aus AMPPS) zu GLMMs
- what is your estimand? paper für das estimand topic

@@

### Define the estimand

#### THEORY

The first step in every research process should always be a clear definition of which theoretical quantity is necessary to answer the specific research question of the study under planning.
Unfortunately, many psychological researchers are still not well trained in this task.
The theoretical estimand is a unit specific quantity defined outside of any statistical model.
Depending on the research and whether the research question is causal, this quantity might be counterfactual and not directly observable.
Part of the theoretical estimand is also a clear definition of the target population, for which the quantity is of interest.

When the theoretical estimand has been specified, the next step is to define the empirical estimand, which reexpresses the theoretical estimand in a way that can be estimated based on the observed data that will be collected in the planned study.
For many common research questions in psychology, the empirical estimand can be expressed as a statistical quantities that can be estimated with a regression model.
Note however, that this is not always possible and that the literature on causal inference includes many estimation strategies beyond regression.

When the empirical estimand has been specified, the last step is to clearly define how a concrete estimate of the empirical estimand shall be computed based on the observed study data.
When a regression model is used for estimation, this includes decisions like which statistical framework and package is used to fit the model, and how to compute hypothesis tests or confidence intervals.

@lundbergWhatYourEstimand2021

#### PRACTICE

In the present case study, we consider the effectiveness of feedback provided by an artificial intelligence (AI) in an AI-enabled diagnostic decision support system.
The context is a clinical setting, where expert radiologists and students under training must detect bleeding based on head scans from computer tomography.
In the investigated AI-enabled diagnostic decision support system, an AI model can provide initial diagnostic advice, which can be used as guidance by the humans require to make the final diagnostic decision.
The research goal is to validate the effectiveness of the AI-enabled advice.
We consider the AI-enable as effective, if the following pattern holds, which we will here describe verbally:

*We expect that for BOTH expert radiologists and students under training, correct AI-advice leads to a higher probability of accurately diagnosing a CT scan compared to no AI-advice presented, AND, we expect that for BOTH experts and non-experts, incorrect advice leads to a lower probability of accurately diagnosing a CT scan compared to no advice presented.*

I becomes clear that our theoretical estimand is be a causal quantity, based on individual quantities define outside of any statistical model.
To explain why, we focus on only on part of the research question concerning the comparison between correct AI-advice and no AI-advice for expert radiologists.
For a specific expert radiologist and a specific CT scan, our question is concerned with the difference between the probability that a correct diagnosis is made if correct AI-advice were presented and the probability that a correct diagnosis is made if no AI-advice were presented.
Mathematically, we could describe this unit-specific quantity with some pseudo-equation like this (where $s$ stands for a specific expert radiologist and $i$ stands for a specific CT scan):

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, a specific expert, a specific scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, a specific expert, a specific scan})
\end{aligned}
$$
However, our verbal description is not precise enough yet, because the above quantity exists and may differ for each combination of expert and scan.
We must specify how we want to aggregate the quantity across different experts and scans.
If we would compute the average of our individual quantity across all experts and scans in the target population, the result would be called the average treatment effect (ATE) in the population of experts.
Although often debated whether this is a good idea, psychological research often focuses on a different quantity.
Instead of averaging over experts and scans, we focus in a hypothetical *typical* expert and a *typical* scan, where *typical* is usually defined as an average score on all attributes of the expert or scan.
With this definition, the theoretical estimand in our above example can be expressed as:

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, an average expert, an average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, an average expert, an average scan})
\end{aligned}
$$
In total, our theoretical estimand consists of four expressions like the one above, the remaining three are:

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, an average expert, an average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, an average expert, an average scan})
\end{aligned}
$$
$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, an average student, an average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, an average student, an average scan})
\end{aligned}
$$

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, an average student, an average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, an average student, an average scan})
\end{aligned}
$$
To complete our definition of the theoretical estimand, we have to clearly define out target population that consists of both persons and scans:
With respect to persons, we are only interested in expert radiologists and medical students at German universities.
With respect to stimuli, we are only interested in the head CT scans made from subjects that do or do not suffer from intracerebral hemorrhage.
Similarly, we are only interested in AI-advice given by the specific AI-enabled diagnostic decision support system under study.

While a theoretical estimand is defined outside of any statistical model and be impossible to observe directly, an empirical estimand is a quantity that can be estimated based on observed data.
For our exemplary research question, it is possible to construct an experimental study, where all participants are confronted with the same set of brain scans, but the kind of AI-advice given for each brain scan is randomly assigned within participants.
This random intervention allows to estimate our theoretical estimand, although in reality, each person receives only one kind of AI-advice (correct advice, wrong advice, no advice) for each brain scan.

For our example, each probability expression in our theoretical estimand can be modeled with a concrete GLMM.
Estimating the GLMM based on the data observed in our planned study produces an estimate for each probability and these estimates can be used to compute an estimate for each of the four probability contrasts.

@@

#### THEORY

When the estimand has been defined, the next step in the research process is to write code that simulates the (assumed) data generating process of the planned study.
This requires specifying a generative process for all predictor variables used in the final data analysis.
While such assumptions can be quite challenging for observational studies and/or continuous predictor variables, this is less of a problem for experimental studies with only categorical predictor variables.
When all predictor variables have been simulated, one can use the structure of a suitable GLMM to simulate the dependent variable.
To simulate the GLMM, one requires plausible values for all model parameters.
We will discuss strategies for how these values can be obtained later.
Because one has full control over the data generating process in a tailored simulation, it is simple to also model highly specific aspects of the planned study, like data missing by design or assuming that subjects dropout (i.e. deleting their data from the simulated dataset) based on other variables in the simulated dataset.
The quality of the results of the sample size planning crucially depends on the plausibility of the simulated data generating process.
For this reason, deciding on a specific set of assumptions can easily feel overwelming.
However, we would argue that even a strongly simplified data generating process (e.g. including only a small number of interaction effect; including only random intercepts but no random slopes; making rather implausible assumptions for missing data) can yield informative result and should be strongly preferred to the current status quo in psychological research, where most studies do not justify their sample size or use highly questionable heuristics that are independent of their specific research question.

#### PRACTICE

We simulate data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system will be evaluated. 
The goal is to understand how AI advice influences medical decision-making.
Participants, radiologists (task experts) and students/interns (non-task experts), review head computer tomography (CT) scans to assess the presence of a bleeding. 
To support their decision-making, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans and if (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the present simulation-based power analysis is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

#### Strategies on specifying the population parameters

#### Insightful descriptive statistics

#### Insightful model based quantities

#### Iterative process with domain experts

@@

@@

@@

```{r}
probs <- avg_predictions(fit, variables = list(
  expert = 0:1, advice_present = 0:1, advice_correct = 0:1), 
  re.form = NA, type = "response")

probs

probs %>% 
  hypotheses(hypothesis = c(
    "b8 - b2 = 0",
    "b7 - b1 = 0",
    "b2 - b4 = 0",
    "b1 - b3 = 0"))
```
